Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-07 Paper/Manuscript_Stats&Comp/reply_letter_STCO.tex
edited rebuttal

\item A safety, general and extensive procedure that allows us to diagnose and automatize the fitting process has been proposed (Section 4).

\item Section 5 has been focused on the main goals of the paper, which are to 1) facilitate the implementation of the method for practical users providing clear formulation of the method and 2) provide a safety and general diagnostic tool to fit and diagnose the model.

\item The analysis has been extended to the Mat\'ern-5/2 kernel. So the whole study finally provides the diagnostics for a squared exponential, Mat\'ern-3/2, Mat\'ern-5/2 (Sections 4.3 and 4.3.1), and periodic squared exponential (Sections 3.3, 4.4 and Appendix C) covariance functions.

@@

\item A diagnostic procedure based on iterative steps to fit and diagnose the approximation has been proposed (Section 4.5.1 and Section 4.6), which is based on: 1) recommendations to set initial values for the parameters, 2) safety diagnostic rules to check if the approximation is sufficiently accurate (Sections 4.5), and 3) numerical equations to automatize updating the parameters in the fitting process (Sections 4.3.1).

@@

\item We have added a new case study to Section 5 which consists in a wiggly 2D simulated data set, demonstrating the applicability and usefulness of the diagnostic tool in multidimensional cases. Additional results concerning computational cost in 2D, 3D and 4D have been presented and briefly discussed, emphasizing that applying the diagnostic tool allows for minimum computational requirements that can be essential for very large data sets, multi-dimensional cases, or other problems that require repeated fitting.

@@

The Hilbert space approximation for GPs (HSGP) is presented as a general formulation for probabilistic modeling, so it can be implemented in different probabilistic programming frameworks, such as Stan, WinBugs, etc. The case study's models have also been formulated in general probabilistic language in the main manuscript, and Stan's implementations have also been provided as examples. Furthermore, HSGP model has been implemented in the \textit{brms} R-package that uses Stan to make model inference, and we are planning to implement the diagnostic tool which consists in iterative steps to fit and diagnose the model also in the \textit{brms} R-package. {\color{blue}(Gabriel: Paul, do we intend to implement the diagnostic procedure in brms?)}

@@

The only limitation of the method is the number of dimensions as these affect computation through the total number of the basis functions needed. But this does not necessarily means a statistical model being simple.
In fact due to this high dependency of the computational requirements of the method on the number of basis functions, the analysis and diagnosis tool developed in this paper arises to be essential for practical applications and in broad scenarios.

@@

It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnosis tool can be essential because it allows iteratively to fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For D>5 even for very smooth functions seems to be impractical.

@@

This work is necessary to facilitate the implementation of the method for practical users, while Solin and Sarka (2020) fundamentally focus on the mathematical development of the model. We also analyze how the key factors of the model relate each other for accurate approximations. We derive numerical functions for these relationships that allow to update the parameters, perform diagnosis, and provide clear and useful guidelines for the whole procedure. So, we make it manageable for practical users, otherwise the practical applicability of the method would be highly limited.

We expect this paper to have a high impact on users as it is highly recommended to make the method usable in practice and in broad scenarios like uni-dimensional functions, multi-dimensional functions, and other problems that require repeated fitting such as cross-validation and simulation based calibration.

We do not see a problem in a purely simulative approach. It simply has different advantages and disadvantages over analytic approaches. In fact, as mentioned above, this study can only be approached from an empirical point of view. We would disagree on the scientific novelty (not all scientific novelty is mathematical in this regard). And we agree with the reviewer about the practical usefulness, which was our goal.

@@

\textbf{It would be really interesting to see a bit more about how HSGP works for functions with 2-3 inputs, and in particular about how a practitioner can recognise whether it makes sense to use HSGP for their case or not. To me it seems clear HSGP is very often a reasonable method for D=1 and almost certainly breaks down for D>=5 or so unless working with very smooth functions (and for them almost any approximation works well), but clear diagnostics for quickly determining whether it could work for D=2 or D=3 would increase the contribution notably. Now the practical process would be to try it out and keep on increasing m until it starts working or one notices the computation is too slow, which feels wasteful -- as a practitioner I would rather just skip HSGP completely if finding out it does not help takes longer than solving the problem without the approximation.}

@@

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already fast. It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnosis tool can be essential because it allows iteratively to fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For D>5 even for very smooth functions seems to be impractical.

@@

Aki: R2 is not happy that we compare only to splines and suggests we should compare to SVI and other variational methods like splines and VI would belong to the same category. We can use Stan's ADVI to run variational inference results. We should include a model for which usual variational inducing point approaches are not directly applicable due to non-exponential family model or combination of different components so that we can argue that we we stick to comparing to black box VI. We have recently run ADVI and improved robust ADVI results for birthday and motorcycle examples (the latter has latent GP for both location and log sigma)
and VI is not that good (it's annoying how common belief it is that black box
VI would be great).

@@

Aki: R2 asks comparisons to Nystrom, but that comparison was
already made by Solin and Särkkä so I don't think we should add that.

>>>>

\item A safe, general, and extensive procedure that allows us to diagnose and automatize the fitting process has been defined explicitly (Section 4).

\item Examples in Section 5 have been clarified.

\item The analysis has been extended to the Mat\'ern-5/2 kernel. So the whole study provides the explicit diagnostic for a squared exponential, Mat\'ern-3/2, Mat\'ern-5/2 (Sections 4.3 and 4.3.1), and periodic squared exponential (Sections 3.3, 4.4 and Appendix C) covariance functions. In addition connection to other covariance functions is discussed.

@@

\item A diagnostic procedure based on iterative steps to fit and diagnose the approximation has been proposed (Section 4.5.1 and Section 4.6), which is based on: 1) recommendations to set initial values for the parameters, 2) diagnostic to check if the approximation is sufficiently accurate (Sections 4.5), and 3) numerical equations to automatize updating the parameters in the fitting process (Sections 4.3.1).

@@

\item We have added a new case study to Section 5 which consists of a wiggly 2D simulated data set, demonstrating the applicability and usefulness of the diagnostic tool in multidimensional cases. Additional results concerning computational cost in 2D, 3D and 4D have been presented and briefly discussed, emphasizing that applying the diagnostic tool allows for minimum computational requirements that can be essential for very large data sets, multi-dimensional cases, or other problems that require repeated fitting.

@@

The Hilbert space approximation for GPs (HSGP) is presented as a general formulation for probabilistic modeling, so it can be implemented in different probabilistic programming frameworks, such as Stan, PyMC, Pyro, Beanmachine, etc. The case study's models have also been formulated in general probabilistic language in the main manuscript, and Stan's implementations have also been provided as examples. Furthermore, HSGP model has been implemented in the \textit{brms} R-package that uses Stan to make model inference, and we are planning to implement the diagnostic tool which consists in iterative steps to fit and diagnose the model also in the \textit{brms} R-package. {\color{blue}(Gabriel: Paul, do we intend to implement the diagnostic procedure in brms?)}

@@

The only limitation of the method is the expoential scaling of the computation with increasing number of dimensions (assuming wigglines stays the same and the amount of data is sufficient to get sufficient informationa about that wiggliness). 
Due to the high dependency of the computational requirements of the method on the number of basis functions, the analysis and diagnosis tool developed in this paper arises to be essential for practical applications and in broad scenarios.

@@

It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnosis tool can be essential because it allows iteratively to fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$ even for very smooth functions seems to be impractical.

@@

Solin and Sarka (2020) fundamentally focus on the mathematical development of the model, and this work is necessary to facilitate the easier use of the method for practical users. We also analyze how the key factors of the model relate each other for accurate approximations. We derive numerical functions for these relationships that allow to update the parameters, perform diagnosis, and provide clear and useful guidelines for the whole procedure. So, we make it manageable for practical users, otherwise the practical applicability of the method would be highly limited.

We expect this paper to have an impact on how widely the basis function GPs are used, as it is makes the method more usable in practice and in broad scenarios like uni-dimensional functions, multi-dimensional functions, and other problems that require repeated fitting such as cross-validation and simulation based calibration.

%We do not see a problem in a purely simulative approach. It simply has different advantages and disadvantages over analytic approaches. In fact, as mentioned above, this study can only be approached from an empirical point of view. We would disagree on the scientific novelty (not all scientific novelty is mathematical in this regard). And we agree with the reviewer about the practical usefulness, which was our goal.

@@

\textbf{It would be really interesting to see a bit more about how HSGP works for functions with 2-3 inputs, and in particular about how a practitioner can recognise whether it makes sense to use HSGP for their case or not. To me it seems clear HSGP is very often a reasonable method for D=1 and almost certainly breaks down for $D>=5$ or so unless working with very smooth functions (and for them almost any approximation works well), but clear diagnostics for quickly determining whether it could work for D=2 or D=3 would increase the contribution notably. Now the practical process would be to try it out and keep on increasing m until it starts working or one notices the computation is too slow, which feels wasteful -- as a practitioner I would rather just skip HSGP completely if finding out it does not help takes longer than solving the problem without the approximation.}

@@

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already fast. It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnosis tool can be essential because it allows iteratively to fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$ even for very smooth functions seems to be impractical.

@@

The main comparison is the Hilbert space basis function approximation against the exact GP as a model. In all comparisons the inference conditional on the model is assumed to be accurate. We've used efficient dynamic HMC, the convergence diagnostics didn't reveal problems, and thus the accuracy of the inference is good. Comparisons of MCMC to variational inference approximations are out of scope of this paper. Some comparisons of MCMC and variational inference for basis function GPs can be found in

Lu Zhang, Bob Carpenter, Andrew Gelman, and Aki Vehtari (2021). Pathfinder: Parallel quasi-Newton variational inference. arXiv preprint arXiv:2108.03782.

Comparisons of the Hilbert space basis function approximations against other GP approximations are provided by Solin and Särkkä (2020), which is cited prominently in our paper.

% Aki: R2 is not happy that we compare only to splines and suggests we should compare to SVI and other variational methods like splines and VI would belong to the same category. We can use Stan's ADVI to run variational inference results. We should include a model for which usual variational inducing point approaches are not directly applicable due to non-exponential family model or combination of different components so that we can argue that we we stick to comparing to black box VI. We have recently run ADVI and improved robust ADVI results for birthday and motorcycle examples (the latter has latent GP for both location and log sigma)
% and VI is not that good (it's annoying how common belief it is that black box
% VI would be great).

@@

Comparisons of the Hilbert space basis function approximations against other GP approximations, including the Nyström method, are provided by Solin and Särkkä (2020), which is cited prominently in our paper.

% Aki: R2 asks comparisons to Nystrom, but that comparison was
% already made by Solin and Särkkä so I don't think we should add that.
